---
title: "Earthquakes Package"
author: "David W. Body"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Earthquakes Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `earthquakes` package consists of four components:

- Data from NOAA Significant Earthquakes Database
- Helper functions to clean the data
- Geoms for plotting earthquake timelines
- Functions for plotting interactive earthquake maps

## Data

Data from the NOAA Significant Earthquakes Database is included with this package.

This is a dataset containing data on destructive earthquakes from 2150 B.C. to the present (mid April 2017). The variables are as follows:

 - I_D
 - FLAG_TSUNAMI
 - YEAR
 - MONTH
 - DAY
 - HOUR
 - MINUTE
 - SECOND
 - FOCAL_DEPTH
 - EQ_PRIMARY
 - EQ_MAG_MW
 - EQ_MAG_MS
 - EQ_MAG_MB
 - EQ_MAG_ML
 - EQ_MAG_MFA
 - EQ_MAG_UNK
 - INTENSITY
 - COUNTRY
 - STATE
 - LOCATION_NAME
 - LATITUDE
 - LONGITUDE
 - REGION_CODE
 - DEATHS
 - DEATHS_DESCRIPTION
 - MISSING
 - MISSING_DESCRIPTION
 - INJURIES
 - INJURIES_DESCRIPTION
 - DAMAGE_MILLIONS_DOLLARS
 - DAMAGE_DESCRIPTION
 - HOUSES_DESTROYED
 - HOUSES_DESTROYED_DESCRIPTION
 - HOUSES_DAMAGED
 - HOUSES_DAMAGED_DESCRIPTION
 - TOTAL_DEATHS
 - TOTAL_DEATHS_DESCRIPTION
 - TOTAL_MISSING
 - TOTAL_MISSING_DESCRIPTION
 - TOTAL_INJURIES
 - TOTAL_INJURIES_DESCRIPTION
 - TOTAL_DAMAGE_MILLIONS_DOLLARS
 - TOTAL_DAMAGE_DESCRIPTION
 - TOTAL_HOUSES_DESTROYED
 - TOTAL_HOUSES_DESTROYED_DESCRIPTION
 - TOTAL_HOUSES_DAMAGED
 - TOTAL_HOUSES_DAMAGED_DESCRIPTION

More information, including variable definitions, is available at https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1

To use the data

```{r, message=FALSE}
filename <- system.file("extdata", "earthquakes.tsv.gz", package = "earthquakes")
earthquakes <- readr::read_delim(filename, delim = "\t")
```

```{r, echo=FALSE}
library(earthquakes)
```

## Functions for cleaning the data

There are two functions for cleaning the earthquakes data.

`eq_clean_data` adds a `DATE` column based on `YEAR`, `MONTH`, `DAY`, converts several character fields to numeric, and cleans `LOCATION_NAME` by passing it to the `eq_location_clean` function.

`eq_location_clean` returns a title case character string with the location's country removed. For example,

```{r}
eq_location_clean("ITALY: VERONA")
```

Create a `clean_earthquakes` data frame like this:

```{r, message=FALSE}
library(dplyr)

clean_earthquakes <- earthquakes %>%
    eq_clean_data()
```

## Geoms for earthquake time lines

This package contains two `ggplot2` geoms for plotting earthquake time lines.

### geom_timeline

`geom_timeline` plots a time line of earthquakes with a point for each earthquake. The `x` aesthetic is a date and the `y` aesthetic is a factor indicating some stratification in which case multiple time lines will be plotted for each level of the factor (e.g. country). Optional aesthetics include `color`, `size`, and `alpha`.

Here is an example:

```{r, fig.show="hold", fig.width=7, message=FALSE}
library(lubridate)
library(ggplot2)

recent_earthquakes <- clean_earthquakes %>%
    filter(COUNTRY == "CHINA", DATE >= ymd('2000-01-01'))#'

g <- ggplot(recent_earthquakes,
            aes(x = DATE, y = COUNTRY, size = EQ_PRIMARY, color = TOTAL_DEATHS))
g <- g + geom_timeline(alpha = 0.5)
g <- g + theme_classic()
g <- g + theme(legend.position = "bottom",
               axis.line.y = element_blank(),
               axis.ticks.y = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank())
g <- g + guides(color = guide_colorbar(title = "# deaths"),
                size = guide_legend("Richter scale value"))
g
```

### geom_timeline_label

This geom is intended to be used in conjunction with the `geom_timeline` geom to add a vertical line with a text annotation (e.g. the location of the earthquake) for each data point on an earthquake timeline. The `x` aesthetic specifies the date of the earthquake and the `label` aesthetic specifies the label for the annotation. An optional `n_max` aesthetic can be used to subset to n_max earthquakes, meaning the n_max largest (by magnitude) will be labeled.

Here is an example:

```{r, fig.width=7, fig.height=7}
recent_earthquakes <- clean_earthquakes %>%
    filter(COUNTRY == "CHINA" | COUNTRY == "USA", DATE >= ymd('2000-01-01'))

g <- ggplot(recent_earthquakes,
            aes(x = DATE, y = COUNTRY, size = EQ_PRIMARY, color = TOTAL_DEATHS))

g <- g + geom_timeline(alpha = 0.5)
g <- g + geom_timeline_label(aes(label = LOCATION_NAME, n_max = 5))
g <- g + theme_classic()
g <- g + theme(legend.position = "bottom",
               axis.line.y = element_blank(),
               axis.ticks.y = element_blank(),
               axis.title.y = element_blank())
g <- g + guides(color = guide_colorbar(title = "# deaths"),
                size = guide_legend("Richter scale value"))
g
```

## Mapping functions

### eq_map

The `eq_map` function #' produces a map of earthquake epicenters (LATITUDE/LONGITUDE) and annotates each point with a popup window containing annotation data stored in a column of the data frame. The radius of each circle is proportional to the earthquake's magnitude (EQ_PRIMARY).

For example:

```{r, fig.width=7, fig.height=7}
clean_earthquakes %>%
    filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>%
    eq_map(annot_col = "DATE")
```

Notice that the date is displayed when you click on an earthquake on the map.

### eq_create_label

This function creates HTML labels with location, magnitude, and total deaths for each earthquake.

Example:

```{r, fig.width=7, fig.height=7}
clean_earthquakes %>%
    filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>%
    dplyr::mutate(popup_text = eq_create_label(.)) %>%
    eq_map(annot_col = "popup_text")
```

Click on earthquakes on the map to see the HTML labels.
